Simple repression draft

### 2020-06-05 doc/appendix_03_bayesian_inference.tex
Not quite final pass on everything, Appdx S3 still needs more

\mmnote{Bayes theorem. Elaborate on posterior as plausibility over space of
model parameters.}

One could argue that the whole goal of formulating theoretical models about
nature is to sharpen our understanding from qualitative statements to precise
quantitative assertions about the relevant features of the natural phenomena in
question \cite{Gunawardena2014}. It is in these models that we intend to distill
the essential parts of the object of study. Writing down such models leads to a
propagation of mathematical variables that parametrize our models. By assigning
numerical values to these parameters we can compute concrete predictions that
can be contrasted with experimental data. But for these predictions to match the
data the parameter values have to carefully be chosen from the whole parameter
space. But how do we go about asserting the effectiveness of different regions
of parameter space to speak to the ability of our model to reproduce the
experimental observations? The language of probability, and more specifically of
Bayesian statistics is -- we think -- the natural language to tackle this
question.

@@

Bayes theorem is a simple mathematical statement that can apply to \textit{any}
logical conjecture. For two particular events $A$ and $B$ that potentially 
depend on each other Bayes theorem gives us a recipe for how to update our 
beliefs about one, let's say $B$, given that we get to observe something about
$A$. In its most classic form Bayes theorem is written as
\begin{equation}
P(B \mid A) = {P(A \mid B) P(B) \over P(A)},
\end{equation}
where the vertical line $\mid$ is read as ``given that''. So $P(B \mid A)$ is
read as probability of $B$ given that $A$ took place. $A$ and $B$ can be any
logical assertion. In particular the problem of Bayesian inference focuses on
the question of finding the probability distribution of a particular parameter
value given the data.

For a given model with a set of parameters $\vec{\theta} = (\theta_1, \theta_2,
\ldots, \theta_n)$, the so-called \textit{posterior distribution} 
$P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
plausibility of a set of parameter values given that we get to observe some
particular dataset. In other words, through the application of Bayes formula we
update our beliefs on the possible values that parameters can take upon learning
the outcome of a particular experiment. We specify the word ``update'' as we
come to every inference problem with prior information about the plausibility of
particular regions of parameter space even before performing any experiment.
Even when we claim as researchers that we are totally ignorant about the values
that the parameters in our models can take, we always come to a problem with
domain expertise that can be exploited. If this was not the case, it is likely
that the formulation of our model is not going to capture the phenomena we claim
to want to understand. This prior information is caputerd in the \textit{prior
probability} $P(\vec{\theta})$. The relationship between how parameter values
can connect with the data is enconded in the \textit{likelihood function} $P(D
\mid \vec{\theta})$. Our theoretical model, whether deterministic or
probabilistic, is encoded in this term that can be intuitively understood as the
probability of having observed the particular experimental data we have at hand
given that our model is parametrized with the concrete values $\vec{\theta}$. 
Implicitly here we are also conditioning on the fact that our theoretical model
is ``true,'' i.e. the model itself if evaluated or simulated in the computer is
capable of generating equivalent datasets to the one we got to observe in an 
experiment. In this way Bayesian inference consists of applying Bayes formula 
as 
\begin{equation}
P(\vec{\theta} \mid D) \propto P(D \mid \vec{\theta}) P(\vec{\theta}).
\end{equation}
Notice than rather than writing the full form of Bayes theorem, we limit 
ourselves to the terms that depend on our quantity of interest -- that is the 
parameter values themselves $\vec{\theta}$ -- as the denominator $P(D)$ only
serves as a normalization constant.

@@

\mmnote{Choosing likelihoods. Easy here, just the steady-state distributions of
our master equations, assuming iid single-cell measurements.}

As we alluded in the previous section it is through the likelihood function 
$P(D \mid \vec{\theta})$ that we encode the connection between our parameter 
values and the experimental observables. Broadly speaking there are two classes
of models that we might need to encode into our likelihood function:
\begin{itemize}
        \item Deterministic models: Models for which a concrete selection of
        parameter values give a single output. Say differently, rigid models 
        with a one-to-one mapping between inputs and outputs.
        \item Probabilistic models: As the name says it, models that rather than
        having a one-to-one mapping, describe the full probability distribution
        of possible outputs.
\end{itemize}
In this paper we focus on inference done with probabilistic models. After all
the chemical master equations we wrote down describe the time evolutions of the
mRNA probability distribution. So all our terms $P(\vec{\theta} \mid D)$ will be
given by the steady-state solution of the corresponding chemical master equation
in question. This is rather convenient as we do not have to worry about adding a
statistical model on top of our model to describe deviations from the
predictions; our models themselves focus on predicting such variations from cell
count to cell count.

\subsubsection{Prior selection}

\mmnote{Choosing priors. Not critical here because we are data-rich and
nonhierarchical, as long as we don't exclude ground truth. In other
applications, prior can matter a lot. This is a feature, not a bug: it's
revealing hidden structure that was implied by your modeling assumptions and you
just didn't realize it. Predictive checks to verify you're not excluding any
plausible values.}

The different models explored in this work embraced different levels of
coarse-graining that resulted in a diverse number of parameters for different
models. For each of these model configurations Bayes theorem demands from us to
represent our preconceptions on the possible parameter values in the form of the
prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is common practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomena in the form of a mathematical model -- which itself is a bold claim of
deeply understanding the features that matter in the object of study -- but we
have absolutely no idea what the parameter values ought to be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down prior distributions for our parameters. In summary then
we don’t want to be maximally uninformative because we aren’t completely
uninformed.

@@

\mmnote{
\begin{itemize}
% \item Bayes theorem.
% Elaborate on posterior as plausibility over
% space of model parameters.
% \item Choosing likelihoods.
% Easy here, just the steady-state distributions of our master
% equations, assuming iid single-cell measurements.
% \item Choosing priors.
% Not critical here because we are data-rich and nonhierarchical,
% as long as we don't exclude ground truth. In other applications,
% prior can matter a lot. This is a feature, not a bug: it's
% revealing hidden structure that was implied by your modeling
% assumptions and you just didn't realize it. Predictive checks to
% verify you're not excluding any plausible values.
\item Marginalization/expectations.
Analytically hard, trivial with sampling.
\item MCMC.
Lot's of fun here, and deep analogies to stat mech and
Hamiltonian mechanics worth exploring. In a perfect world, you'd
wave a magic wand and generate independent draws from target
posterior. But we don't know how. Next best: generate a Markov
chain of $N$ \textit{correlated} samples, estimate an
autocorrelation time $\tau$, then you've got roughly $N/\tau$
effective samples. Obviously with some assumptions.
\item Posterior predictive checks.
So the posterior is fine, but is the model actually not wrong?
PPC is essential to verify that your model is a plausible
generative process for the data.
\item Discuss identifiability, degeneracy, and failed PPCs in main text. Better thru example anyway.
\end{itemize}
}

@@

What this means is that we have access to data with different combinations of
$k^-$ and $k^+$. We could naively try to fit the kinetic parameters individually
for each of the datasets, but there is no reason to believe that the binding
site identity for the LacI repressor somehow affects its expression level
controlled from a completely different location in the genome, nor vice versa.
In other words, what makes the most sense it to fit all datasets together to
obtain a single value for each of the association and dissociation rates. What
this means, as described in~\ref{section_04_bayesian_inference} of the main text
is that we have a seven dimensional parameter space with four possible
association rates $k^+$ given the four available aTc concentrations, and three
possible dissociation rates $k^-$ given the three different binding sites
available in the dataset.

@@

\mrm{This paragraph needs its own subsection in my opinion if it is to be 
included in the SI.}
This likelihood is rather inscrutable. We did not find any of the known
analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
we instead resorted to numerics. One insight we found
\mmnote{a plot would probably help explain, maybe SI, need to explain better}
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
values, except with $k_i$ multiplied by the fold-change
$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the
\textit{ratio} $k_R^+/k_R^-$ was detectable. But for intermediate repression,
the distribution was visibly broadened with Fano factor greater than $1+b$, the
value for the corresponding constitutive case. This indicates that the repressor
rates had left an imprint on the distribution, and perhaps intuitively, this
intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
$k_i$. Put another way, if the repressor rates are much faster or much slower
than $k_i$, then there is a timescale separation and effectively only one
timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
rates in the problem are comparable does the mRNA distribution retain detectable
information about them.

@@

\mmnote{Outline:
\begin{itemize}
\item Hey look it mostly works. With full model, 9D model is well identified,
though most of the individual experiments unsurprisingly are not:
only 3 of the 9 datasets were identifiable if fit in isolation
and for the rest only the ratio of repressor rates is identifiable.
Only by making assumptions that certain rates
are equal across datasets was identifiability achieved.
PPC is not perfect, it kinda misses some
rep/op pairs but overall it's surprisingly predictive considering how strong
were the assumptions that went into it.
\item Do we chalk this up to experimental imperfections or is there interesting
biology to be found in the disagreements? I'm inclined towards the former. Even
just catching cells in truly apples-to-apples growth phases, or matching up
parameters of imaging sessions weeks or months apart, is hardly trivial.
\item The alternative is to refine our model and/or add on complexity, and it's
not at all clear what additions to make to the model, especially how to produce
the long tail of repressed strains over UV5. I see no way to get that without
getting into complicated time-correlation/history effects, where
transcription...rebounds?? After being repressed?? Sounds odd.
\item Comparison between my inferred rates, equilibrium binding E, and
single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
really! With exception of one Oid data data point, free energies are w/in about
0.5 kT also. That's not massively larger than the repeatability threshold of
fold-change measurements, at maybe 0.1 to 0.3 kT.
\item Surprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
\end{itemize}
}

>>>>

One could argue that the whole goal of formulating theoretical models about
nature is to sharpen our understanding from qualitative statements to precise
quantitative assertions about the relevant features of the natural phenomena in
question \cite{Gunawardena2014}. It is in these models that we intend to distill
the essential parts of the object of study. Writing down such models leads to a
propagation of mathematical variables that parametrize our models. By assigning
numerical values to these parameters we can compute concrete predictions that
can be contrasted with experimental data. For these predictions to match the
data the parameter values have to carefully be chosen from the whole parameter
space. But how do we go about assessing the effectiveness of different regions
of parameter space to speak to the ability of our model to reproduce the
experimental observations? The language of probability, and more specifically of
Bayesian statistics is -- we think -- the natural language to tackle this
question.

@@

Bayes theorem is a simple mathematical statement that can apply to \textit{any}
logical conjecture. For two particular events $A$ and $B$ that potentially 
depend on each other Bayes theorem gives us a recipe for how to update our 
beliefs about one, let us say $B$, given some state of knowledge, or lack thereof, about
$A$. In its most classic form Bayes theorem is written as
\begin{equation}
P(B \mid A) = {P(A \mid B) P(B) \over P(A)},
\end{equation}
where the vertical line $\mid$ is read as ``given that''. So $P(B \mid A)$ is
read as probability of $B$ given that $A$ took place. $A$ and $B$ can be any
logical assertion. In particular the problem of Bayesian inference focuses on
the question of finding the probability distribution of a particular parameter
value given the data.

For a given model with a set of parameters $\vec{\theta} = (\theta_1, \theta_2,
\ldots, \theta_n)$, the so-called \textit{posterior distribution} 
$P(\vec{\theta} \mid D)$, where $D$ is the experimental data, quantifies the
plausibility of a set of parameter values given our observation of some
particular dataset. In other words, through the application of Bayes formula we
update our beliefs on the possible values that parameters can take upon learning
the outcome of a particular experiment. We specify the word ``update'' as we
come to every inference problem with prior information about the plausibility of
particular regions of parameter space even before performing any experiment.
Even when we claim as researchers that we are totally ignorant about the values
that the parameters in our models can take, we always come to a problem with
domain expertise that can be exploited. If this was not the case, it is likely
that the formulation of our model is not going to capture the phenomena we claim
to want to understand. This prior information is captured in the \textit{prior
probability} $P(\vec{\theta})$. The relationship between how parameter values
can connect with the data is enconded in the \textit{likelihood function} $P(D
\mid \vec{\theta})$. Our theoretical model, whether deterministic or
probabilistic, is encoded in this term that can be intuitively understood as the
probability of having observed the particular experimental data we have at hand
given that our model is parametrized with the concrete values $\vec{\theta}$. 
Implicitly here we are also conditioning on the fact that our theoretical model
is ``true,'' i.e. the model itself if evaluated or simulated in the computer is
capable of generating equivalent datasets to the one we got to observe in an 
experiment. In this way Bayesian inference consists of applying Bayes formula 
as 
\begin{equation}
P(\vec{\theta} \mid D) \propto P(D \mid \vec{\theta}) P(\vec{\theta}).
\end{equation}
Notice than rather than writing the full form of Bayes theorem, we limit 
ourselves to the terms that depend on our quantity of interest -- that is the 
parameter values themselves $\vec{\theta}$ -- as the denominator $P(D)$ only
serves as a normalization constant.

We also emphasize that the dichotomy we have presented between prior and
likelihood is more subtle. Although it is often stated that our prior
knowledge is entirely encapsulated by the obviously named prior
probability $P(\vec{\theta})$, this is usually too simplistic.
The form(s) we choose for our likelihood function
$P(D \mid \vec{\theta})$ also draw heavily on our prior domain expertise
and the assumptions, implicit and explicit, that these choices encode are
at least as important, and often inseparable from,
the prior probability, as persuasively argued in~\cite{Gelman2017}.

@@

As we alluded in the previous section it is through the likelihood function 
$P(D \mid \vec{\theta})$ that we encode the connection between our parameter 
values and the experimental observables. Broadly speaking there are two classes
of models that we might need to encode into our likelihood function:
\begin{itemize}
        \item Deterministic models: Models for which a concrete selection of
        parameter values give a single output. Said differently, models 
        with a one-to-one mapping between inputs and outputs.
        \item Probabilistic models: As the name suggests, models that, rather than
        having a one-to-one input-output mapping, describe the full
        probability distribution of possible outputs.
\end{itemize}
In this paper we focus on inference done with probabilistic models. After all,
the chemical master equations we wrote down describe the time evolutions of the
mRNA probability distribution. So all our terms $P(\vec{\theta} \mid D)$ will be
given by the steady-state solution of the corresponding chemical master equation
in question. This is rather convenient as we do not have to worry about adding a
statistical model on top of our model to describe deviations from the
predictions. Instead our models themselves focus on predicting such variation
in cell count.

\subsubsection{Prior selection}
The different models explored in this work embraced different levels of
coarse-graining that resulted in a diverse number of parameters for different
models. For each of these model configurations Bayes theorem demands from us to
represent our preconceptions on the possible parameter values in the form of the
prior $P(\vec{\theta})$. Throughout this work for models with $> 1$ parameter we
assign independent priors to each of the parameters; this is
\begin{equation}
P(\vec{\theta}) = \prod_{i=1}^n P(\theta_i).
\end{equation}
Although it is common practice to use non-informative, or maximally
uninformative priors, we are of the mindset that this is a disservice to the
philosophical and practical implications of Bayes theorem. It sounds almost
contradictory to claim that can we represent our thinking about a natural
phenomena in the form of a mathematical model -- which itself is a bold claim of
deeply understanding the features that matter in the object of study -- but we
have absolutely no idea what the parameter values ought to be. We therefore make
use of our own expertise, many times in the form of order-of-magnitude
estimates, to write down prior distributions for our parameters. In summary then
we don’t want to be maximally uninformative because we aren’t completely
uninformed.

@@

\subsubsection{Expectations and marginalizations}
One of the great powers of Bayesian analysis is \textit{marginalization}.
For models with more than one or two parameters, it is generally unwise
to attempt to draw inferences directly from the full joint posterior
distribution $P(\vec{\theta} \mid D)$...

\mmnote{
\begin{itemize}
% \item Bayes theorem.
% Elaborate on posterior as plausibility over
% space of model parameters.
% \item Choosing likelihoods.
% Easy here, just the steady-state distributions of our master
% equations, assuming iid single-cell measurements.
% \item Choosing priors.
% Not critical here because we are data-rich and nonhierarchical,
% as long as we don't exclude ground truth. In other applications,
% prior can matter a lot. This is a feature, not a bug: it's
% revealing hidden structure that was implied by your modeling
% assumptions and you just didn't realize it. Predictive checks to
% verify you're not excluding any plausible values.
\item Marginalization/expectations.
Analytically hard, trivial with sampling.
\item MCMC.
Lot's of fun and deep analogies to stat mech and
Hamiltonian mechanics we'll allude to but not explore here.
The idea of MCMC: in a perfect world, you'd
wave a magic wand and generate independent draws from target
posterior. But we don't know how. Next best: generate a Markov
chain of $N$ \textit{correlated} samples, estimate an
autocorrelation time $\tau$, then you've got roughly $N/\tau$
effective samples. Obviously with some assumptions.
\item Posterior predictive checks.
So the posterior is fine, but is the model actually not wrong?
PPC is essential to verify that your model is
not blatantly contradicted by the data.
\end{itemize}
}

@@

What this means is that we have access to data with different combinations of
$k^-$ and $k^+$. We could naively try to fit the kinetic parameters individually
for each of the datasets, but there is no reason to believe that the binding
site identity for the LacI repressor somehow affects its expression level
controlled from a completely different location in the genome, nor vice versa.
In other words, what makes the most sense it to fit all datasets together to
obtain a single value for each of the association and dissociation rates. What
this means, as described in Section~\ref{section_04_bayesian_inference} of the main text
is that we have a seven dimensional parameter space with four possible
association rates $k^+$ given the four available aTc concentrations, and three
possible dissociation rates $k^-$ given the three different binding sites
available in the dataset.

@@

@@

\mmnote{Outline:
Perhaps urprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
}

### 2020-06-05 doc/section_01_introduction.tex
Not quite final pass on everything, Appdx S3 still needs more

Gene expression presides over much of the most important dynamism of living
organisms. The level of expression of batteries of different genes is altered as
a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals. The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}. One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems first
and foremost from our ignorance about what those architectures even are with
more than 60\% of the genes even in an ostensibly well understood organism such
as {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites of
a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. One root to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Garcia2011, Phillips2019}. If we demonstrate that we can pass that
test by successfully predicting both the means and variance in gene expression,
then that provides a more solid foundation upon which to launch into more
complex problems.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on the quantitative
dissection of transcription \cite{Oehler1994, Oehler1990}. All of the proposed
models models coarse-grain away some of the important microscopic features of
this architecture that have been elucidated by generations of geneticists,
molecular biologists and biochemists. Our aim is to understand the underlying
principles of such coarse-graining in order to increase the predictive power of
these models. In doing so we hope future models of regulatory response will be
able to serve the powerful predictive role needed to take synthetic biology from
a brilliant exercise in enlightened empiricism to a rational design framework as
in any other branch of engineering. More precisely, we want phenomenology in the
sense of coarse-graining away atomistic detail, but still retaining biophysical
meaning. For example, we are not satisfied with the strictly phenomenological
approach offered by the commonly used Hill functions~\cite{Frank2013}. Studies
like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are clearly
insufficient since each new situation requires a completely new set of
parameters. Such work requires a quantitative theory of how biophysical changes
at the molecular level propagate to input-output functions at the genetic
circuit level. We want concepts, not mere facts. In particular a key question
is: at this level of coarse-graining, what microscopic details do we need to
explicitly model, and how do we figure that out? For example, do we need to
worry about all or even any of the steps that individual RNA polymerases go
through each time they make a transcript? Turning the question around, can we
see any imprint of those processes in the available data? If the answer is no,
then those processes are irrelevant for our purposes. Forward modeling and
inverse (statistical inferential) modeling are necessary to tackles such
questions.

>>>>

Gene expression presides over much of the most important dynamism of living
organisms. The level of expression of batteries of different genes is altered as
a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals. The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}. One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems first
and foremost from our ignorance about what those architectures even are with
more than 60\% of the genes even in an ostensibly well understood organism such
as {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites of
a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Gertz2009, Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}. One route to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Garcia2011, Phillips2019}. If we demonstrate that we can pass that
test by successfully predicting both the means and variance in gene expression,
then that provides a more solid foundation upon which to launch into more
complex problems.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site as developed in pioneering early work on the quantitative
dissection of transcription \cite{Oehler1994, Oehler1990}. All of the proposed
models coarse-grain away some of the important microscopic features of
this architecture that have been elucidated by generations of geneticists,
molecular biologists and biochemists. Our aim is to understand the underlying
principles of such coarse-graining in order to increase the predictive power of
these models. In doing so, we hope future models of regulatory response will be
able to serve the powerful predictive role needed to take synthetic biology from
a brilliant exercise in enlightened empiricism to a rational design framework as
in any other branch of engineering. More precisely, we want phenomenology in the
sense of coarse-graining away atomistic detail, but still retaining biophysical
meaning. For example, we are not satisfied with the strictly phenomenological
approach offered by the commonly used Hill functions. As argued
in~\cite{Frank2013}, Hill functions are ubiquitous precisely because they
coarse-grain away all biophysical details into inscrutable parameters. Studies
like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are clearly
insufficient since each new situation requires a completely new set of
parameters. Such work requires a quantitative theory of how biophysical changes
at the molecular level propagate to input-output functions at the genetic
circuit level. We want concepts, not mere facts. In particular a key question
is: at this level of coarse-graining, what microscopic details do we need to
explicitly model, and how do we figure that out? For example, do we need to
worry about all or even any of the steps that individual RNA polymerases go
through each time they make a transcript? Turning the question around, can we
see any imprint of those processes in the available data? If the answer is no,
then those processes are irrelevant for our purposes. Forward modeling and
inverse (statistical inferential) modeling are necessary to tackle such
questions.
